\documentclass{article}
  \usepackage{fullpage}
  \usepackage{amssymb}
  \usepackage{amsthm}
  \usepackage{graphicx}
  \usepackage{url}
  \usepackage{pdfpages}

\begin{document}
\author{Derek Seiple}
\title{How to Add a Test tpl to the ADMB Test Folder}
\date{}
\maketitle

\section{Set-up}
This guide will show you how to add a test tpl to the test folder in ADMB.
For our purposes we will assume we have the following tpl that we want to add
\begin{verbatim}
solve.tpl
\end{verbatim}
along with all of its supporting files. In this example we only have one
supporting file, \verb"solve.dat".

In ADMB the test folder is located in the main parent directory. We want to
make sure we are in that folder; that is maneuver to
\begin{verbatim}
~/admb/test$
\end{verbatim}
where \verb"admb" is the name of the parent ADMB directory.

First, we create a folder for our tpl and supporting files (\verb"RC04\_solve" in this
example). You can do this however is comfortable for you. To do it by command line type
\begin{verbatim}
~/admb/test$ mkdir RC04_solve
\end{verbatim}
Next, move the tpl file as well as any supporting files into the folder we just
created.

\section{Create the GNUmakefile}

At this point we will need a make file. Inside the \verb"RC04_solve" folder
create a file called \verb"GNUmakefile". Now you will have to edit that file and put
the following in it (notice that make files use tabs).

\begin{verbatim}
.PHONY: run

TARGET=solve

all: clean $(TARGET) run

$(TARGET): $(TARGET).tpl $(TARGET).dat
        export ADMB_HOME=$(ADMB_HOME); export PATH=$(ADMB_HOME)/bin:${PATH}; admb -s $(TARGET)

run:
        @printf "Started run:\n"
        ./$(TARGET)
        @printf "Finished run.\n"

clean:
        @rm -vf $(TARGET)
        @rm -vf admodel.*
        @rm -vf variance
        @rm -vf fmin.log
        @rm -vf $(TARGET).eva
        @rm -vf $(TARGET).htp
        @rm -vf $(TARGET).bar
        @rm -vf $(TARGET).bgs
        @rm -vf $(TARGET).cor
        @rm -vf $(TARGET).cpp
        @rm -vf $(TARGET).log
        @rm -vf $(TARGET).o
        @rm -vf $(TARGET).par
        @rm -vf eigv.rpt
\end{verbatim}

The majority of this file will not change for most of the test tpls that you
will add. But there are a couple things that will change.

The line that contains \verb"TARGET=solve" will change depending on the name of the tpl.
If your tpl file is \verb"your_tpl_file.tpl", then that line will read:
\begin{verbatim}
TARGET=your_tpl_file
\end{verbatim}

Also, if your tpl has random effects in it, then the end of line 8 will be
\begin{verbatim}
admb -r -s $(TARGET)
\end{verbatim}
instead of
\begin{verbatim}
admb -s $(TARGET)
\end{verbatim}

Lastly, you may need to add some more lines at the end of the \verb"clean:" section
depending on your tpl file. If you run \verb"make" followed by \verb"make clean" and
some extra files are there that were not before, then add those file names to the end.
For example if \verb"extra.file" is left over then add
\begin{verbatim}
@rm -vf extra.file
\end{verbatim}
to the end of the clean section.

\section{Edit Makefile}

The last thing you will have to do is add this example to the main Makefile in
\begin{verbatim}
~/admb/test$
\end{verbatim}

When you open \verb"~/admb/test/Makefile" it will look something like this, but with a few more entries:
\begin{verbatim}
.PHONY: clean testminmax admb_messages 

all: testminmax admb_messages

testminmax:
        OPTIONS=-s make --directory=$@

admb_messages:
        OPTIONS=-s make --directory=$@

clean:
        make --directory=testminmax clean
        make --directory=admb_messages clean
\end{verbatim}

We will make the following changes:

On the first line (contains \verb".PHONY:") we will add \verb"RC04_solve" (the name of the
folder we created earlier) at the end.

On the line with the \verb"all:" tag we will add \verb"RC04_solve" at the end.

Right before we get to the \verb"clean:" tag we will put the following:
\begin{verbatim}
RC04_solve:
        OPTIONS=-s make --directory=$@
\end{verbatim}

Lastly, we will add
\begin{verbatim}
make --directory=RC04_solve clean
\end{verbatim}
to the end of the \verb"clean:" section.

So when it is all said and done the Makefile should look something like this:
\begin{verbatim}
.PHONY: clean testminmax admb_messages RC04_solve

all: testminmax admb_messages RC04_solve

testminmax:
        OPTIONS=-s make --directory=$@

admb_messages:
        OPTIONS=-s make --directory=$@

RC04_solve:
        OPTIONS=-s make --directory=$@

clean:
        make --directory=testminmax clean
        make --directory=admb_messages clean
        make --directory=RC04_solve clean
\end{verbatim}



\end{document}
